---
title: "Recreating Gapminder"
output: github_document
---

```{r message = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(gapminder)
library(maps)
```

## The Visualization

```{r cache = TRUE}
ggplot(data = gapminder) +
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp, 
                           size = pop, color = continent), alpha = 0.5) +
  labs(x = "Income", 
       y = "Life Expectancy", 
       color = "World Region", 
       size = "Population", 
       title = "Gapminder Bubble Chart", 
       subtitle = "For all years")
```


## More with dplyr

### `filter()`, `select()`, and `arrange()`

```{r}
gapminder
```

Create a copy of gapminder.

```{r}
my_gm <- gapminder
my_gm
```

View only USA

```{r}
my_gm %>% 
  filter(country == "United States")
```

Plot `gdpPercap` over time (`year`) for only USA

```{r}
my_gm %>% 
  filter(country == "United States") %>% 
  ggplot(aes(x = year, y = gdpPercap)) +
  geom_line()
```

## May 12

Subset of gapminder with only USA

```{r}
us_gm <- my_gm %>% 
  filter(country == "United States")
```

View only year, life expectancy, and gdp percap only for USA

```{r}
us_gm %>% 
  select(year, lifeExp, gdpPercap) %>% 
  arrange(desc(lifeExp))
```

Reorgaize US subset with life expectancy first and drop continent

```{r}
us_gm %>% 
  select(lifeExp, everything(), -continent)
```

rename variables

```{r}
us_gm %>% 
  select(life_exp = lifeExp, everything())
```

Show the entries for Burundi after 1996 for only the variables `yr`, `life_exp`, and `pop`.


```{r}
my_gm %>% 
  filter(country == "Burundi" & year > 1996) %>% 
  select(yr = year, life_exp = lifeExp, pop)
```
## May 14

Create `gdp` variable

```{r}
my_gm %>% 
  mutate(gdp = gdpPercap * pop,
         gdp_billion = gdp / 1000000000)
```

```{r}
my_gm %>% 
  mutate(gdp = gdpPercap * pop,
         gdp_billion = gdp / 1000000000,
         gdp = NULL)
```

What does this code do?

```{r}
us_tib <-  my_gm %>%                                         # copy my_gm into us_tib object, then
  filter(country == "United States")                         # keep only the United States values
## This is a semi-dangerous way to do this variable
## I'd prefer to join on year, 
## but we haven't covered joins yet (but will next week!)
my_gm <-  my_gm %>%                                          # copy my_gm into my_gm object, then
  mutate(tmp = rep(us_tib$gdpPercap, nlevels(country)),      # create variable tmp, that is                                                                        repeating US gdpPercap for how many                                                                 countries there are
         gdpPercap_rel_US = gdpPercap / tmp,                # create variable var that is the ratio                                                                of gdpPercap for that country                                                                       compared to gdpPercap of US
         tmp = NULL)                                         # remove tmp variable
```

Is the US a high GDP country?

```{r}
my_gm %>% 
  ggplot(aes(x = gdpPercap_rel_US)) +
  geom_histogram()
```


| Function type        | Explanation | Examples | In `dplyr` |
|----------------------|-------------|----------|------------|
| Vectorized functions | These take a vector, and operate on each component to return a vector of the same length (i.e., element-wise). | `cos`, `sin`, `log`, `exp`, `round` | `mutate` |
| Aggregate functions  | These take a vector, and return a vector of length 1 | `mean`, `sd`, `length`, `typeof` | `summarize` - can be in combination with `group_by`. |
| Window functions     | these take a vector, and return a vector of the same length that depends on the vector as a whole. | `lag`, `rank`, `cumsum` | `mutate` in combination `group_by` |

Three ways to count in R (`dplyr`)

How many values/observations are in each continent?

```{r}
my_gm %>% 
  group_by(continent) %>% 
  summarize(n = n())
```

```{r}
my_gm %>% 
  group_by(continent) %>% 
  tally()
```

```{r}
my_gm %>% 
  count(continent)
```

```{r}
my_gm %>% 
  group_by(continent) %>% 
  summarize(n = n(),
            n_countries = n_distinct(country))
```

Compute summaries for multiple variables.

Compute the average and median `lifeExp` and `gdpPercap` by `continent` and `year`, but only for 1952 and 2007.

```{r}
my_gm %>% 
  filter(year %in% c(1952, 2007)) %>% 
  group_by(continent, year) %>% 
  summarize_at(vars(lifeExp, gdpPercap),
               list(mean = mean,
                    median = median))
```

versus no names in variables (not preferred):

```{r}
my_gm %>% 
  filter(year %in% c(1952, 2007)) %>% 
  group_by(continent, year) %>% 
  summarize_at(vars(lifeExp, gdpPercap),
               list(mean, median))
```

Grouped mutates
View growth in population since first year of record by each country

```{r}
my_gm %>% 
  group_by(country) %>% 
  select(country, year, pop) %>% 
  mutate(pop_gain = pop - first(pop)) %>% 
  filter(year < 1963)
```

Calculate growth in life expectancy for each country compared to 1972 for each country.

```{r}
my_gm %>% 
  group_by(country) %>% 
  select(country, year, lifeExp) %>% 
  mutate(life_exp_gain = lifeExp - nth(lifeExp, n = 5))
```

```{r}
my_gm %>%
  filter(continent == "Asia") %>%
  select(year, country, lifeExp) %>%
  group_by(year) %>%
  filter(min_rank(desc(lifeExp)) < 2 | min_rank(lifeExp) < 2) %>% 
  arrange(year) %>%
  print(n = Inf)
```

Challenge:
Which five countries had sharpest 5-year drop in lifeExp?

```{r}
my_gm %>% 
  group_by(country) %>% 
  mutate(change_life_exp = lifeExp - lag(lifeExp, n = 1)) %>% 
  ungroup() %>% 
  top_n(n = 5, wt = change_life_exp) %>% 
  select(country, continent, year, change_life_exp) %>% 
  arrange(desc(change_life_exp))
```

What explains these drops in life expectancy? Post in community/Issue!


### May 21 Session

```{r}
gm_long <- gapminder %>% 
  pivot_longer(lifeExp:gdpPercap,
               names_to = "measure",
               values_to = "amount")
gm_long
```

```{r}
gm_wide <- gapminder %>% 
  pivot_wider(names_from = year,
              values_from = c(lifeExp, pop, gdpPercap))
gm_wide
```

### wide to long with only gdp

Using `pivot_longer()` and `separate()`.

```{r}
gm_wide %>%
  select(country, continent, starts_with("gdp")) %>%
  pivot_longer(starts_with("gdp"),
               names_to = "year",
               values_to = "gdppercap") %>%
  separate(year, into = c(NA, "year"), sep = "_")
```

Using `pivot_longer()` and `names_sep = ...`

```{r}
gm_wide %>% 
  select(country, continent, starts_with("gdp")) %>%
  pivot_longer(starts_with("gdp"),
               names_to = c(NA, "year"),
               values_to = "gdpPercap",
               names_sep = 10)
```

Using `pivot_longer()` and `names_pattern = "..."`

```{r}
gm_wide %>% 
  select(country, continent, starts_with("gdp")) %>% 
  pivot_longer(starts_with("gdp"),
               names_to = "year",
               values_to = "gdpPercap",
               names_pattern = "gdpPercap_(.*)")
```

### long to wide

```{r}
gm_long %>% 
  pivot_wider(names_from = c(year, measure),
              values_from = amount,
              names_prefix = "silly_")
```

Some text examples:

silly-1952-lifeExp - nope (double-clicking is awkward)
silly:1952:lifeExp - nope (double-clicking is awkward)
silly_1952_lifeExp - preferred

## May 28

Coloring a map of the world


```{r world-map, cache = TRUE}
ggplot(map_data("world"), aes(x = long, y = lat)) +
  geom_polygon(aes(group = group))

# ggplot(map_data("world2"), aes(x = long, y = lat)) +
#   geom_polygon(aes(group = group))
```


Adding color

```{r color-map, cache = TRUE}
ggplot(map_data("world"), aes(x = long, y = lat)) +
  geom_polygon(aes(group = group, fill = group))
```

Combining shape file with continent information

```{r joining-shape-continent, message = FALSE}
world_shapes <- map_data("world") %>% as_tibble()

world_continent <- world_shapes %>% 
  left_join(gapminder %>% select(country, continent),
            by = c("region" = "country"))
```


Plot world, color by continent

```{r color-continent, cache = TRUE}
ggplot(world_continent, aes(x = long, y = lat)) +
  geom_polygon(aes(group = group, fill = continent))
```

```{r}
world_continent %>% 
  filter(is.na(continent)) %>% 
  count(continent)
```


```{r}
anti_join(gapminder %>% select(country, continent),
          world_shapes %>% select(region),
          by = c("country" = "region")) %>% 
  count(country)
```

```{r}
anti_join(world_shapes %>% select(region),
          gapminder %>% select(country, continent),
          by = c("region" = "country")) %>% 
  count(region)
```

Relabel gapminder countries

```{r}
gm_region <- gapminder %>% 
  mutate(
    country = as.character(country),
    region = case_when(
      country == "Congo, Dem. Rep." ~ "Democratic Republic of the Congo",
      country == "Congo, Rep." ~ "Republic of Congo",
      country == "Cote d'Ivoire" ~ "Ivory Coast",
      country == "Korea, Dem. Rep." ~ "North Korea",
      country == "Korea, Rep." ~ "South Korea",
      country == "Slovak Republic" ~ "Slovakia",
      country == "Trinidad and Tobago" ~ "Trinidad",
      country == "United Kingdom" ~ "UK",
      country == "United States" ~ "USA",
      country == "West Bank and Gaza" ~ "Palestine",
      country == "Yemen, Rep." ~ "Yemen",
      TRUE ~ country
  ))
```


Join relabeled values to shape file

```{r}
world_continent <- left_join(world_shapes,
                             gm_region,
                             by = "region")
```


Map relabeled continents

```{r cache = TRUE}
ggplot(world_continent, aes(x = long, y = lat)) +
  geom_polygon(aes(group = group, fill = continent))
```

Life expectancy in 2007 across the globe

```{r cache = TRUE}
world_continent %>% 
  filter(year == 2007) %>% 
  ggplot(aes(x = long, y = lat)) +
  geom_polygon(aes(group = group, fill = lifeExp))
```


gdpPercap in 2007 across the globe


```{r cache = TRUE}
world_continent %>% 
  filter(year == 2007) %>% 
  ggplot(aes(x = long, y = lat)) +
  geom_polygon(aes(group = group, fill = gdpPercap))
```

## June 2

Factors

```{r inspect-factors}
str(gapminder$continent)
levels(gapminder$continent)
nlevels(gapminder$continent)
class(gapminder$continent)

gapminder %>% 
  count(continent)
```

### Unused factors

```{r}
nlevels(gapminder$country)

set.seed(123)
samp_countries <- sample(unique(gapminder$country), size = 7)

samp_gm <- gapminder %>% 
  filter(country %in% samp_countries)

nlevels(samp_gm$country)

samp_gm_drop <- samp_gm %>% 
  droplevels()

nlevels(samp_gm_drop$country)
```

### Change factor ordering

By default, countries/continent are ordered alphabetically

```{r}
gapminder %>% 
  ggplot(aes(y = continent)) +
  geom_bar()
```

#### By frequency

```{r}
gapminder %>% 
  mutate(continent = fct_infreq(continent)) %>% 
  ggplot(aes(y = continent)) +
  geom_bar()

gapminder %>% 
  ggplot(aes(y = fct_infreq(continent))) +
  geom_bar() +
  labs(y = "continent")
```

#### By reverse freq

```{r}
gapminder %>% 
  mutate(continent = fct_infreq(continent),
         continent = fct_rev(continent)) %>% 
  ggplot(aes(y = continent)) +
  geom_bar()
```

#### By another variable

Plot European countries ordered by lifeExp

```{r}
gm_europe_2007 <- gapminder %>% 
  filter(year == 2007, continent == "Europe")

gm_europe_2007 %>% 
  ggplot(aes(x = lifeExp, y = country)) +
  geom_point()

gm_europe_2007 %>% 
  ggplot(aes(x = lifeExp, y = fct_reorder(country, lifeExp))) +
  geom_point() +
  labs(y = "country")
```

#### By two variables

```{r}
samp_gm %>% 
  ggplot(aes(x = year, y = lifeExp, color = country)) +
  geom_line()

samp_gm %>% 
  ggplot(aes(x = year, y = lifeExp,
                  color = fct_reorder2(country, year, lifeExp))) +
  geom_line() +
  labs(color = "country") # to make better looking label
```

### Recoding levels

```{r}
nlevels(gapminder$country)

three_gm <- gapminder %>% 
  filter(country %in% c("United States", "Greece", "Australia")) %>% 
  droplevels()

nlevels(three_gm$country)

three_gm$country %>% levels()

three_gm$country %>% 
  fct_recode("US" = "United States", "Oz" = "Australia") %>% 
  levels()

three_gm2 <- three_gm %>% 
  mutate(country = fct_recode(country, "US" = "United States", "Oz" = "Australia"))

```

### Combining factors

```{r}
gm1 <- gapminder %>% 
  filter(country %in% c("United States", "Mexico"), year > 2000) %>% 
  droplevels()

gm2 <- gapminder %>% 
  filter(country %in% c("France", "Germany"), year > 2000) %>% 
  droplevels()

levels(gm1$country)
levels(gm2$country)
```

Can we simply combine these datasets?

```{r}
c(gm1$country,gm2$country)

fct_c(gm1$country,gm2$country)
```












